NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-01-13 main.tex
Updates from Overleaf

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. the importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, as well as for nucleic acids~\cite{hospital16}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}) and arbitrary biomolecules~\cite{bekker20}. However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access is still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

@@

\subsection{Programmatic access to MD simulation data of membranes composed of most common biologically abundant lipids}

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. Nearly thousand simulation trajectories with the total length approaching half a millisecond can be accessed programmatically using the Python application programming interface (API) or a graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}). Following the principles of improving data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} and NMRlipids project protocol, not only the content and computer programs related to the databank, but also the entire construction process of the databank are open access~\cite{botan15}. 
%The NMRlipids databank will be always open for data and other contributions.

Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. The distribution of available simulations containing a specific lipid is shown in Fig.~\ref{fig:overlay}C. This roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. The force fields used in the simulations cover all the essential parameters commonly used in lipid simulations, ranging from highly popular CHARMM36 parameters to united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

The efficient upcycling of large MD simulation trajectories in the NMRlipids databank is enabled by the overlay structure illustrated in Fig.~\ref{fig:overlay}A. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability and permanent links to the data, such as digital object identifiers (DOIs) or uniform resource names (URNs). The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored in the git repository which is currently available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}) and results from analyses described in sections below. A more detailed description on the NMRlipids databank structure is available the supplementary information.

@@

@@

\subsection{Analyzing the databank}
Programs that analyze large data sets from the NMRlipids databank loop over README.yaml files of all simulations, download the raw simulation data to the local computer, and perform the analysis from the downloaded trajectories. When molecule or atom names are needed in the analysis, they will be read from the README.yaml file and mapping files defined therein.

>>>>

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. the importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access is still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

@@

\subsection{Programmatic access to MD simulations of membranes composed of most common biologically abundant lipids}

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} using the NMRlipids project protocol where all the content is openly accessible from the beginning of the project~\cite{botan15}.  The NMRlipids databank contains currently nearly thousand simulation trajectories with the total length approaching half a millisecond. The distribution of available simulations containing a specific lipid in Fig.~\ref{fig:overlay}C roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}.Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}E, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

The overlay structure, illustrated in Fig.~\ref{fig:overlay}A, is designed to enable efficient upcycling of MD simulations for data-driven and machine learning applications using 
The NMRlipids databank. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability, such as Zenodo (\url{zenodo.org}), and with permanent links to the data, such as digital object identifiers (DOIs). 
%or uniform resource names (URNs). 
The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored in the git repository which is available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}) and results from analyses described in sections below which are available at \url{https://github.com/NMRLipids/DataBankManuscript}. A more detailed description on the NMRlipids databank structure is available the supplementary information.

@@

\subsection{Accessing NMRlipids databank}
The content of the NMRlipids databank is easily searchable and browsable via graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}), which can be also used to visualize the systems, compare basic properties between diffent systems and experiments, order the simulations based on their quality.

%\subsection{Analyzing the databank}
More flexible analyses, enabling wide range of novel applications can be performed by the Python application programming interface (API) available at \url{https://github.com/NMRLipids/Databank}. 
Programs that analyze large data sets from the NMRlipids databank loop over README.yaml files of all simulations, download the raw simulation data to the local computer, and perform the analysis from the downloaded trajectories. When molecule or atom names are needed in the analysis, they will be read from the README.yaml file and mapping files defined therein.

@@
